---
title: "1. Interactive data visualisation with plotly and R"
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    self_contained: yes
    toc: false
    fig_width: 10
    fig_height: 8
    depth: 3  # upto three depths of headings (specified by #, ## and ###)  

vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r}
suppressPackageStartupMessages({
  library("datasauRus")
  library("plotly")
  library("magrittr")
  library("plotlyutils")
  library("tidyr")
})
```

# Background

- Talk is available as an R package on github (or contact the author).

# Motivation - why do we plot data?

- Summarise
- Explore
- Discover

# Anscombes quartet
```{r}
data(anscombe)
df <- gather(anscombe) %>%
  mutate(dataset = gsub("[xy](\\d)", "\\1", key)) %>%
  mutate(key = gsub("\\d", "", key)) %>%
  group_by(key) %>% 
  mutate(id=1:n()) %>% 
  spread(key, value) %>% 
  select(-id)

lms <- df %>% 
  group_by(dataset) %>% 
  summarise(
    intercept = broom::tidy(lm(y ~ x))[1, "estimate"][[1]],
    slope = broom::tidy(lm(y ~ x))[2, "estimate"][[1]],
    x = min(df$x),
    xend = max(df$x),
    y = (x * slope) + intercept,
    yend = (xend * slope) + intercept
  )


ax <- list(
  title = "",
  zeroline = FALSE,
  showline = FALSE,
  showticklabels = FALSE,
  showgrid = FALSE
)

plot_ly(df, 
    x = ~x, 
    y = ~y, 
    frame = ~dataset, 
    mode = "markers",
    type = "scatter",
    showlegend = FALSE) %>%
  add_segments(
    data = lms,
    inherit = FALSE,
    x = ~x,
    xend = ~xend,
    y = ~y,
    yend = ~yend,
    frame = ~dataset
  ) %>%
    layout(xaxis = ax, yaxis = ax) %>%
    animation_opts(frame = 2500, transition = 500)

```

# Dirty Dozen

```{r}

datasaurus_dozen$dataset %<>% factor(
    levels = c(
        "away", 
        "high_lines", 
        "wide_lines",
        "h_lines", 
        "v_lines",
        "slant_down", 
        "slant_up", 
        "dots",
        "bullseye",
        "circle",
        "star", 
        "x_shape",
        "dino")
    )

plot_ly(datasaurus_dozen, 
    x = ~x, 
    y = ~y, 
    frame = ~dataset, 
    mode = "markers",
    type = "scatter",
    showlegend = FALSE) %>%
    layout(xaxis = ax, yaxis = ax) %>%
    animation_opts(frame = 2500, transition = 500)

```


# What can we do with plotly?

- ggplotly - convert static ggplot objects to dynamic graphs
- Scatter plots
- Barplots
- Boxplots
- Line plots (including density plots)
- Heatmaps
- Combinations of all of the above!
- Custom `htmlwidgets` with useful functionality

# Not covered

- Shiny
